---
title: "CarterSun_A2"
author: "Carter Sun"
date: "10/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
library(sf)
library(tigris)
library(mapview)
library(leaflet)
library(censusapi)
Sys.setenv(CENSUS_KEY="ae50ac1baf05a1c9f156a1c83c62722ac7838e04")
```

```{r include=FALSE}
dec_vars_2020 <-
  listCensusMetadata(
    name = "2020/dec/pl",
    type = "variables"
  )

smc_pop_2020 <-
  getCensus(
    name = "dec/pl",
    vintage = 2020,
    region = "block:*",
    regionin = "state:06+county:081",
    vars = "P1_001N"
  ) %>% 
  transmute(
    block = 
      paste0(state,county,tract,block),
    pop = P1_001N
  )

blocks_2020 <-
  blocks(
    "CA", "San Mateo", year = 2020, progress_bar = F
    )

```

```{r include=FALSE}
dec_vars_2010 <-
  listCensusMetadata(
    name = "2010/dec/pl",
    type = "variables"
  )

smc_pop_2010 <-
  getCensus(
    name = "dec/pl",
    vintage = 2010,
    region = "block:*",
    regionin = "state:06+county:081",
    vars = "P001001"
  ) %>% 
  transmute(
    block = 
      paste0(state,county,tract,block),
    pop = P001001
  )

blocks_2010 <-
  blocks(
    "CA", "San Mateo", year = 2010, progress_bar = F
    )
```



```{r include=FALSE}

nfo_boundary <- places("CA", progress_bar = F) %>% 
  filter(NAME == "North Fair Oaks")

nfo_blocks_2010 <- blocks_2010 %>% 
  st_centroid() %>% 
  .[nfo_boundary, ] %>% 
  st_drop_geometry() %>% 
  left_join(blocks_2010 %>% select(GEOID10)) %>% 
  st_as_sf() %>% 
  select(block = GEOID10) %>% 
  left_join(smc_pop_2010)

```


```{r include=FALSE}
nfo_blocks_2020 <- blocks_2020 %>% 
  .[nfo_boundary, ] %>% 
  select(block = GEOID20) %>% 
  left_join(smc_pop_2020)
```

```{r include=FALSE}
nfo_blocks_2020_intersect <- 
  nfo_blocks_2020 %>% 
  st_transform(26910) %>% 
  mutate(original_area = st_area(.)) %>% 
  st_intersection(
    nfo_blocks_2010 %>% 
      select(block) %>% 
      st_transform(26910)
  ) %>% 
  mutate(
    leftover_area = st_area(.),
    perc_area = as.numeric(leftover_area/original_area),
    pop_2020 = pop * perc_area 
  )
```


```{r include=FALSE}
nfo_blocks_2020_reshaped <- 
  nfo_blocks_2020_intersect %>%
  st_drop_geometry() %>% 
  group_by(block) %>% 
  summarize(
    pop_2020 = sum(pop_2020, na.rm=T) %>% round()
  )
```

```{r include=FALSE}
projection <- "+proj=longlat +zone=10 +ellps=GRS80  +datum=WGS84 +units=ft +no_defs"
#+datum=NAD83 +proj=utm
nfo_pop_2010_2020 <- nfo_blocks_2010 %>% 
 # st_drop_geometry() %>% 
  left_join(nfo_blocks_2020_reshaped) %>% 
#  st_as_sf() %>% 
  mutate(
    absolute_pop = (pop_2020 - pop)/ as.numeric(st_area(.)) * 43560
  )

```


```{r include=FALSE}
absolute_pal <- colorNumeric(
  palette = "PiYG",
  domain = c(-500,600)
)
```

```{r echo=FALSE}
leaflet(nfo_pop_2010_2020) %>% 
  addProviderTiles(provider = providers$CartoDB.Positron) %>% 
  addPolygons(
    fillColor = ~absolute_pal(absolute_pop),
    label = ~paste0(
      round(absolute_pop), 
      " people per acre"
      ),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.75,
    weight = 1
  ) %>% 
  addLegend(
    pal = absolute_pal,
    value = c(-500,600),
    title = "Absolute Change <br> in Density"
  )
```

The findings in the map show that there was very little population change (less than 150) in North Fair Oaks from 2010 to 2020. North Fair Oaks is a lower-income neighborhood, so the small change in density makes sense, as people who are looking to move into the Bay Area might opt for a wealthier area, and people who currently live there might not have the financial resources to move somewhere more expensive, with the Bay Area already being one of the most expensive places to live. An interesting observation I made was that a lot of the areas blocks along the train tracks seemed to be rapidly increasing in population. This might be because houses along the train tracks are cheaper, since there is more noise from the train, allowing low-income families to afford cheaper housing. One caveat to the map is the grayed areas, which are grey for two reasons. One reason is is because they are outliers in the data, and don't fit in the -400 to 600 range, which was arbitrarily chosen as it seemingly best represented the data. The second reason is that the absolute change in density was N/A. This happens for a variety of reasons, one of which being that it is a commercial zone, as seen in the leftmost block along El Camino Real, which is a Target and therefore has nobody living there.
