---
title: "CarterSun_A4"
author: "Carter Sun"
date: "10/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE)
```

```{r include=FALSE}
library(tidyverse)
library(censusapi)

Sys.setenv(CENSUS_KEY="ae50ac1baf05a1c9f156a1c83c62722ac7838e04")

acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )
```


```{r include=FALSE}
race_categories <- c(
  "White alone", "Black or African American alone", "American Indian and Alaska Native alone", "Asian alone", "Native Hawaiian and Other Pacific Islander alone", "Some other race alone", "Two or more races"
)

bay_education_race <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = paste0("group(C15002",LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA,"education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>%   
      group_by(county, education) %>% 
      summarize(estimate = sum(estimate)) %>% 
      mutate(race = race_categories[x]) 
      
  })
bay_education_race <- subset(bay_education_race,bay_education_race$county == "081")
      
```

```{r echo=FALSE}
bay_race_total <-
  bay_education_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(education = "Total")

bay_education_race %>% 
  group_by(education, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(bay_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = education %>% factor(levels = rev(c("Total","Less than high school diploma","High school graduate (includes equivalency)","Some college or associate's degree", "Bachelor's degree or higher"))),
      y = estimate,
      fill = race %>% factor(levels = rev(unique(bay_education_race$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Education Attainment",
    y = "Proportion of population 25 years or older",
    title = "San Mateo County Education\nAttainment by Race",
    fill = "Race of householder"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```

```{r include=FALSE}
acs_vars_2019_1yr <-
  listCensusMetadata(
    name = "2019/acs/acs1",
    type = "variables"
  )


ethnicity_categories <- c(
  "White, not Latinx","Latinx"
)

bay_education_ethnicity <-
  8:9 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs1",
      vintage = 2019,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = paste0("group(B15002",LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_1yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,NA,"education"),
        sep = "!!"
      ) %>% 
      filter(!is.na(education)) %>%   
      group_by(county, education) %>% 
      summarize(estimate = sum(estimate)) %>% 
      mutate(ethnicity = ethnicity_categories[x-7]) 
  }) %>% 
  subset(county == "081") %>% 
  mutate (
    education = 
    case_when(
      education %in% c("9th to 12th grade, no diploma", "Less than 9th grade") ~ "Less than high school diploma",
      education %in% c("Regular high school diploma", "GED or alternative credential") ~ "High school graduate (includes equivalency)",
      education %in% c("Associate's degree", "Some college, no degree") ~ "Some college or associate's degree",
      education %in% c("Bachelor's degree", "Graduate or professional degree") ~ "Bachelor's degree or higher"
    )
  ) %>% 
  group_by(county, ethnicity, education) %>% 
  summarize(estimate = sum(estimate))
  
    
```


```{r include=FALSE}
total <- getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = "group(B15002)"
  ) %>%
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "name",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_5yr %>% 
      select(name, label)
  ) %>% 
  select(-name) %>% 
  separate(
    label,
    into = c(NA,NA,NA,"education"),
    sep = "!!"
  ) %>% 
  filter(!is.na(education)) %>% 
  group_by(county, education) %>% 
  summarize(estimate = sum(estimate)) %>% 
  subset(county == "081")%>% 
  mutate (
    education = 
    case_when(
      education %in% c("No schooling completed","Nursery to 4th grade","5th and 6th grade","7th and 8th grade","9th grade","10th grade", "11th grade", "12th grade, no diploma") ~ "Less than high school diploma",
      education %in% c("High school graduate (includes equivalency)") ~ "High school graduate (includes equivalency)",
      education %in% c("Associate's degree", "Some college, less than 1 year", "Some college, 1 or more years, no degree") ~ "Some college or associate's degree",
      education %in% c("Bachelor's degree", "Doctorate degree", "Master's degree", "Professional school degree") ~ "Bachelor's degree or higher"
    )
  ) %>% 
  group_by(county, education) %>% 
  summarize(estimate = sum(estimate))


total_white_latinx <-
  bay_education_ethnicity %>% 
  group_by(county, education) %>% 
  summarize(estimate = -1*sum(estimate))

not_white_not_latinx <-
  rbind(total, total_white_latinx) %>% 
  group_by(county, education) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(ethnicity = "Not white, not Latinx")

bay_education_ethnicity <-
  rbind(bay_education_ethnicity, not_white_not_latinx)
```

```{r echo=FALSE}
bay_ethnicity_total <-
  bay_education_ethnicity %>% 
  group_by(ethnicity) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(education = "Total")

bay_education_ethnicity %>% 
  group_by(education, ethnicity) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(bay_ethnicity_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = education %>% factor(levels = rev(c("Total","Less than high school diploma","High school graduate (includes equivalency)","Some college or associate's degree", "Bachelor's degree or higher"))),
      y = estimate,
      fill = ethnicity %>% factor(levels = rev(unique(bay_education_ethnicity$ethnicity)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Education Attainment",
    y = "Proportion of population 25 years or older",
    title = "San Mateo County Education\nAttainment by Ethnicity",
    fill = "Ethnicity of householder"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```

In San Mateo county, there is definitely a disporportionate education attainment by both race and ethnicity. For African Americans, there is a considerably large proportion of their population that only has a high school diploma or a GED, while the proportion of African Americans with a Bachelor's degree or higher is a lot smaller. However, for white householders, who are statistically more affluent than African Americans in San Mateo County, are more evenly spread, with the highest proportion having a Bachelor's degree or higher. Similarly, while looking at ethnicity, the majority Latinx housholders have either less than or a high school diploma, while a small proportion has any college experience at all. Inversely, the largest proportion of White, not Latinx housholders have a Bachelor's degree or higher, while the smallest has less than a high school diploma. This definitely points to a clear disproportion in education attainment in San Mateo County.


