---
title: "CarterSun_A8"
author: "Carter Sun"
date: "11/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r include=FALSE}
library(tigris)
library(tidyverse)
library(tidycensus)
library(sf)
library(censusapi)
library(leaflet)
library(StatMatch)

Sys.setenv(CENSUS_KEY="ae50ac1baf05a1c9f156a1c83c62722ac7838e04")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()
```
```{r echo=FALSE, warning=FALSE}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_pumas,
    weight = 1,
    color = "gray",
    label = ~PUMACE10
  ) %>% 
  addMarkers(
    lng = -121.7825387,
    lat = 37.99537
  ) %>% 
  addPolygons(
    data = bay_pumas %>% 
      filter(PUMACE10 == "01309")
  )
```
```{r include=FALSE}
pums_bart <-
  readRDS("pums_2014_2019_wts.rds") %>% 
  mutate(
    PWGTP = as.numeric(PWGTP),
    bart = ifelse(
      JWTR %in% c("4"),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA, year) %>% 
  summarize(
    pop = sum(PWGTP),
    bart = sum(bart)
  )
```
```{r echo=FALSE}
pums_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = pums_bart %>% 
    filter(year == 2018) %>% 
    pull(pop)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2018) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(pop),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": Population ", pop)
  )
```
```{r echo=FALSE}
pums_pal <- colorNumeric(
  palette = "GnBu",
  domain = pums_bart %>% 
    filter(year == 2018) %>% 
    pull(bart)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2018) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(bart),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": ", bart, " BART commute riders")
  )
```

```{r include=FALSE}
pums_bart_clean <-
  pums_bart %>% 
  select(-pop) %>% 
  pivot_wider(
    names_from = year,
    values_from = bart
  )
```

```{r include=FALSE}
obs_matrix <-
  pums_bart_clean %>% 
  ungroup() %>% 
  select(`2014`,`2015`,`2016`) %>% 
  as.matrix()

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- pums_bart_clean$PUMA
colnames(dist_matrix) <- pums_bart_clean$PUMA

match <- dist_matrix["01309",] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    PUMA = rowname,
    match = "."
  ) %>% 
  right_join(
    pums_bart_clean
  ) %>% 
  arrange(match) %>% 
  .[1:11, ] %>% 
  left_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
  st_as_sf()
```
```{r echo=FALSE}
leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match[1, ],
    color = "red",
    label = ~PUMA
  ) %>% 
  addPolygons(
    data = match[-1, ],
    label = ~PUMA
  )
```
```{r echo=FALSE}
match_pumas <-
  match %>% 
  filter(!PUMA %in% c("01309")) %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>%
  group_by(
    year
  ) %>% 
  summarize(
    bart = mean(bart),
    PUMA = "Similar PUMAs"
  )

treatment_pumas <-
  match %>% 
  filter(PUMA %in% c("01309")) %>% 
  select(-match) %>% 
  st_drop_geometry() %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  )

rbind(
  treatment_pumas,
  match_pumas
) %>% 
  ggplot(
    aes(
      x = as.numeric(year),
      y = bart,
      color = PUMA
    )
  ) +
  geom_line() +
  geom_vline(xintercept = 2018, linetype = "dashed") +
  labs(
    title = "Antioch vs. control neighborhoods, BART ridership",
    x = "Year",
    y = "BART commute riders"
  )
```
```{r echo=FALSE}
transit_did <-
  match %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>% 
  mutate(
    year = year %>% as.numeric(),
    time = ifelse(year >= 2018, 1, 0),
    treated = ifelse(PUMA == "01309", 1, 0)
  )

did_reg <- lm(bart ~ treated*time, data = transit_did)

summary(did_reg)
```
The model shows the Antioch BART station in 2018 having an estimated impact of 967 new BART commuters. This result is statistically significant, because along with the positive effect size of the treatment and the line graph, the p value is 2.109e^-6, or 0.52%. However, one thing to note is that from 2017 to 2018, the number of riders in the 01309 PUMA was already rising dramatically, and after 2018, the number plateaued, while the overall number of BART riders was riding. Furthermore, the station opened in mid-2018, and we don't know when the survey was taken. The overall increase of BART riders also might be due to the opening of a BART station in Milpitas in Spring of 2017. Milpitas is also in one of the matched PUMAS, so it might be affecting the difference in differences estimate. Another assumption I made is that I arbitrarily chose ten matching PUMAs using Mahalanobis distance. However, we don't know if there were more or fewer PUMAs, so ten might be the wrong number to use. Another assumption comes directly from the ACS survey, where our data comes from. Question 32 asks about how the responder commutes to work, and the options are subway or elevated rail, long-distance train or commuter rail, and light rail, streetcar, or trolley, to name a few of the relevant options. I assumed that all the people who rode the BART, and only the people who rode the BART, selected subway or elevated rail. However, people who ride the BART may have selected some of the other options like commuter rail, or some people who don't ride the BART, but other forms of public transit like the MUNI, might have selected subway or elevated rail. If the number of people that did this is substantial enough, it might have had an effect on my model and difference in differences estimate. For a difference in differences analysis, the quality of the model significantly depends on the judgment, assumptions, and biases of the person designing the analysis. All in all, however, because there was such a large positive effect size of the treatment and in the line graph, as well as a very low p-value, the model shows that the Antioch BART station opening led to more people to use BART to commute to work.